### 2020-07-27 TCS_v1/01-introduction.tex
pulizia e riorganizzazione

Then, we present a methodology for verifying formally
properties of Reaction Systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
where one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to consider for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the Reaction Systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant introduced here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

>>>>

Then, we present a methodology for verifying formally
properties of Reaction Systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
when one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to consider for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the Reaction Systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant introduced here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

### 2020-07-27 TCS_v1/05-translation.tex
pulizia e riorganizzazione

\[
\begin{array}{lcl@{\hskip 1cm}lcl}
 P_s &\defeq & E(s,\widetilde{s},\widehat{s},\underline{s}) 
 &
 \overline{P_s}& \defeq & E(\overline{s},\widetilde{s},\widehat{s},\underline{s})
\end{array}
\]
%{\color{red}
%\[
%\begin{array}{rcl}
% P(s,\tilde{s},\hat{s},\underline{s}) & \defeq &\sum_{h_j,k_j\in\{0,1\}}
%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^{h_j} \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^{k_j}\ \link{\hat{s}_i}{\hat{s}_o}.P_s\\
%&&\ + \\
%&&
%\sum_{\scriptsize \begin{array}{l}h_j,k_j, \\ h'_j\in\{0,1\}\end{array}}
%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^{h_j} \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^{k_j} \  \startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact} \    (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^{h'_j} \ \link{\underline{s}_i}{\underline{s}_o}.P_s \\
%
%%\sum_{R_1+R_2\geq 0}(\sum_{h,k \in[0,1]} 
%%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^k)^{R_1} \startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact} (\sum_{h ,k\in[0,1]}
%%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^k)^{R_2} \link{\underline{s}_i}{\underline{s}_o}.P_s\\
%&&\ +\\
%&&\sum_{j\geq 0}(
%  \startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^j\ \ \link{\underline{s}_i}{\underline{s}_o}.\overline{P_s}
%\end{array}
%\]
%}
\[
\begin{array}{rcl}
 E(s,\widetilde{s},\widehat{s},\underline{s}) & \defeq &\sum_{h,k\geq 0}  
  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ \startchain{\widehat{s}_i}\chainedlink{\widehat{s}_o}{\noact}\chainend{\noact} (\startchain{\widetilde{s}_i}\chainedlink{\widetilde{s}_o}{\noact}\chainend{\noact})^k.P_s\\
&&\ + \\
&&
\sum_{h\geq 0,k\geq 1}
 (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ \startchain{\underline{s}_i}\chainedlink{\underline{s}_o}{\noact}\chainend{\noact}(\startchain{\widetilde{s}_i}\chainedlink{\widetilde{s}_o}{\noact}\chainend{\noact})^k.P_s\\
&&\ +\\
&&\sum_{h\geq 0}(\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ \link{\underline{s}_i}{\underline{s}_o}.\overline{P_s}
\end{array}
\]
The first line of $ E(s,\widetilde{s},\widehat{s},\underline{s})$ accounts for the case where $s$ is tested for presence by $h$ reactions and produced by $k$ reactions, while being provided by the context ($\link{\widehat{s}_i}{\widehat{s}_o}$).
Thus, $s$ will be present at the next step (the continuation is $P_s$). Here $h$ and $k$ are not known a priori and therefore any
combination is possible. By knowing the number of reactions that test $s$, we can bound the maximum 
values of $h$ and $k$.
The second line accounts for the analogous case where $s$ is not provided by the context ($\link{\underline{s}_i}{\underline{s}_o}$). 
The condition $k\geq 1$ guarantees that $s$ will remain present (the continuation is $P_s$).
%Here taking
%$h = \sum_{j\geq0} h_j +h'_j$, $k = (\sum_{j \geq 0} k_j )  + 1$ guarantees that $s$ will be present in the next step. Please note that 
The third line accounts for the case where $s$ is tested for presence, but it is neither produced nor provided by  the context. Therefore, in the next step $s$ will be absent in the system (the continuation is $\overline{P_s}$).
Note that in the case of $\overline{P_s}$ the test for presence of $s$ in the system is just replaced by the test for its absence.
%Process $P_s$ (and $\overline{P_s}$) offers link chain prefixes for all the possible
% permutations of the solid links; nevertheless if the reactions, and their enumeration, is known
% the code of $P_s$ can be drastically simplified.

@@

\noindent
We exploit names $r_j,p_j$ to join the chains provided by the application of all the reactions. 
Channels $r_j$ and $r_{j+1}$ enclose the enabling/disabling condition of reaction $aj$.
Channels $p_i$ and $p_{j+1}$ enclose the links related to the entities produced by $aj$.
We will see that all the link chain labels of transitions follow the same schema: first we find all the reactions limited to the reactants and inhibitors (chained using $r_j$ channels), then all the supplies by the contexts (chained using the channel $\mathit{cxt}$, to be introduced next), and finally the products for all the reactions (chained using $p_j$ channels). 
In the following there is an example explaining this schema.
%\[
%\begin{array}{lcl@{\qquad}l}
%P_{aj}& \defeq & 
% ^{r_j} \backslash  \blockchain{s_i}{s_o}{s \in R_j} \, \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \in I_j} \, \link{}{}
% \blockchain{\tilde{c}_i}{\tilde{c}_o}{c \in P_j}\, \backslash_{r_{j+1}}.P_{aj} & \mbox{\small  \{$aj$ is applicable\}}\\
%&& + \ \\
%&& \sum_{s\in R_j}\startchain{r_j}\chainedlink{\overline{s}_i}{\noact}\chainedlink{\noact}{\overline{s}_o} \chainend{r_{j+1}}.P_{aj}  & \mbox{\small  \{$aj$ is not applicable\}}\\
%&&+\\
%&& \sum_{e\in I_j}\startchain{r_j}\chainedlink{e_i}{\noact} \chainedlink{\noact}{e_o} \chainend{r_{i+1}}.P_{aj}  & \mbox{\small \{$aj$ is not applicable\}}\\
%\end{array}
%\]

\paragraph{Processes for contexts}
The context can vary at each step $n$ and for each entity $s \in S$, the context must say if the entity is provided or not. Let us denote by $C_n$ the context at step $n$, i.e. the set of entities that are made available. Correspondingly, we introduce another process $\mathit{Cxt}^n$ defined as follows:
\[
\begin{array}{lcl}
\mathit{Cxt}^n &\defeq &
%\left \{
^{cxt} \backslash  \blockchain{s_i}{s_o}{s \in C_n} \, \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \not\in C_n} \,  \backslash_{p_1}. \mathit{Cxt}^{n+1}
%+ \\
% ^{cxt} \backslash  \blockchain{s_i}{s_o}{s \in I2_n} \, \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \in Oj_n} \, \backslash_{p_1}
% \chainedlink{\hat{s}_i}{\noact}\chainedlink{\noact}{\hat{s}_o}\chainend{p_1}. Cxt_{s}^{n+1} & \ \mbox{if the context provides $s$  at the $n$-th step}\\[5pt] 
% \startchain{cxt_j}\chainedlink{\underline{s}_i}{\noact}\chainedlink{\noact}{\underline{s}_o}\chainend{cxt_{j+1}}.Cxt_s^{n+1} & \ \mbox{otherwise}
%\end{array}\right.\\
%Cxt_s & \defeq & Cxt_s^1\\
%\end{array}
%\right.\\
\end{array}
\]

@@

\begin{example}\label{ex:backbone}
 Let ${\cal A}$ be a RS whose specification contains two entities, $s1$ and $s2$, and the reactions $r_1=(s1,\ , s2)$ and $r_2 = (s2, \ , s1)$ that produce $s2$ if $s1$ is present and $s1$ if $s2$ is present. For simplicity we do not consider inhibitors, but the reader can assume a void inhibitor is present in both reactions. Then, we assume an  extended interactive process $\pi=(\gamma,\delta)$ where the context $\gamma$ provides  $s1$ and $s2$ at every step, but we assume that only $s1$ is initially present. The corresponding c\CNA \ process is $\llbracket{\cal A},\gamma\rrbracket \defeq \restrict{\mathit{names}} \llparenthesis{\cal A},\gamma\rrparenthesis$, with 
$$
\llparenthesis{\cal A},\gamma\rrparenthesis \defeq 
I ~|~ P_{s1}~|~\overline{P}_{s2}~|~ P_{r1}~|~P_{r2}~|~\mathit{Cxt}
$$
 where:
 \[
 \begin{array}{rcl}
 P_{r_1} &\defeq& \startchain{r_1}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}\chainedlink{r_2}{\noact} \chainedlink{\noact}{p_1}\chainedlink{\widetilde{s2}_i}{\noact}\chainedlink{\noact}{\widetilde{s2}_o}\chainend{p_2}.P_{r_1} + \dots;   \\[6pt]
  P_{r_2} &\defeq& \dots + \startchain{r_2}\chainedlink{\overline{s2}_i}{\noact}\chainedlink{\noact}{\overline{s2}_o}\chainedlink{\mathit{cxt}}{\noact} \chainedlink{\noact}{p_2}\chainend{\silent}.P_{r_2} + \dots; \\[6pt]
   P_{s1} &\defeq& \startchain{s1_i}\chainedlink{s1_o}{\noact}\chainedlink{\noact}{\widehat{s1}_i}\chainend{\widehat{s1}_o}
 %{\noact}\chainedlink{\noact}{\tilde{s1}_i}\chainedlink{\tilde{s1}_o}
 .P_{s1} + \dots; \\[6pt]
  \overline{P}_{s2} &\defeq& \startchain{\overline{s2}_i}\chainedlink{\overline{s2}_o}{\noact}
   \chainedlink{\noact} {\widehat{s2}_i}\chainedlink{\widehat{s2}_o}{\noact}\chainedlink{\noact} {\widetilde{s2}_i}\chainend{\widetilde{s2}_o}.P_{s2} + \dots;\\[6pt]
 \mathit{Cxt}& \defeq & \startchain{\mathit{cxt}}\chainedlink{\widehat{s1}_i}{\noact}\chainedlink{\noact}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\noact}\chainedlink{\noact}{\widehat{s2}_o}\chainend{p_1}.\mathit{Cxt}
  %Cxt& \defeq & \startchain{cxt}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}
  %\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}.Cxt
 %Cxt_{s1}& \defeq & \startchain{cxt_1}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}\chainend{cxt_2}.Cxt_{s1} &
 % Cxt _{s2} &\defeq& \startchain{cxt_2}\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}. Cxt_{s2}
\end{array}
\]  \\
For clarity of exposition, we show the code of the processes just in part, to focus on the prefixes that will be involved in the first transition of the system. In Figure~\ref{fig:backbone} we show the  structure of a link chain label related to the execution of such a transition. The yellow blocks are referred to init process $I$, to the processes encoding the reactions, $P_{r_1}$ and $P_{r_2}$, and to the context $\mathit{Cxt}$. As the figure puts in evidence, these two kinds of processes determine the structure of the link chain, from end to end, i.e. from the left $\silent$ to the right one. We could say that these processes form the \emph{backbone} of the interaction.
   In contrast, the processes encoding the entities, $P_{s1}$,  $\overline{P}_{s2}$, provide the
    solid links to be merged with the virtual links of the backbone (i.e. to be plugged in the backbone).
 Formally, we have
 $$
 \llparenthesis{\cal A},\gamma\rrparenthesis \xrightarrow{
 \startchain{\silent}
 \chainedlink{r_1}{r_1}
 \chainedlink{s1_i}{s1_i}
 \chainedlink{s1_o}{s1_o}
 \chainedlink{r_2}{r_2}
 \chainedlink{\overline{s2}_i}{\overline{s2}_i}
 \chainedlink{\overline{s2}_o}{\overline{s2}_o}
 \chainedlink{\mathit{cxt}}{\mathit{cxt}}
 \chainedlink{\widehat{s1}_i}{\widehat{s1}_i}
 \chainedlink{\widehat{s1}_o}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\widehat{s2}_i}
  \chainedlink{\widehat{s2}_o}{\widehat{s2}_o}
  \chainedlink{p_1}{p_1}
  \chainedlink{\widetilde{s2}_i}{\widetilde{s2}_i}
  \chainedlink{\widetilde{s2}_o}{\widetilde{s2}_o}
  \chainedlink{p_2}{p_2}
  \chainend{\silent}
 }
 P'
 $$

@@

\paragraph{The flat function}
For technical reasons, our transition labels are quite verbose; then, to simplify their processing, we introduce a function that takes a solid link chain and returns a simple string by eliminating all the channel matching pairs leaving just one placeholder for them. This transformation is harmless, in the sense that it retains all the information in the chain, because it is applied to complete chains only. 
The function $\mathit{flat}$ is defined inductively as follows:
%\[
%\begin{array}{lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl}
%\mathit{flat}(\epsilon)&\defeq & \epsilon &
%\mathit{flat}(\link{\gamma}{\gamma'}) &  \defeq & \left \{ \begin{array}{cl} \epsilon & \mathrm{if } (\gamma = \beta_i \wedge \gamma = \beta_o)  \vee \gamma = \silent\\
%\gamma' & \mathrm{otw }  \end{array} \right. &
%%\mathit{flat}(\link{\alpha}{\silent})& \defeq & \epsilon &
%%\mathit{flat}(\link{\alpha}{\beta}) & \defeq &\beta \\
%\end{array}\]

@@

% \begin{proposition}[correctness2]
% \label{prop:correctness1}
% Let $P$ be a c\CNA \ process such that $\exists$ a $rs$ ${\cal A}$ and a $n$-step interactive process $\pi=(\gamma,\delta)$ with $\gamma =C_0,\dots,C_n$ and $\delta = D_0,\dots,D_n$ such that $P= [|{\cal A},\gamma|]$. 
% %We set $\pi'=(\gamma',\delta')$ with $\gamma'=C_0',\dots,C_{n-1}'$,  $\delta'=D_0',\dots,D_{n-1}$,  where $C_0'= C_1 \cup D_1$, $C_i'= C_{i+1}$, $D_i=D_{i+1}'$,  and $1 \leq i < n$.
%Then,   $\exists$ $P'$ and $\pi'=(\gamma',\delta')$ such that $P \xrightarrow{\link{\silent}{\silent}}P'$, $P' =[|{\cal A},\gamma'|]$.
% \end{proposition} 
\noindent
Now, we extend the previous result to a series of transitions.

>>>>

\[
\begin{array}{lcl@{\hskip 1cm}lcl}
 P_s &\defeq & E(s,\widetilde{s},\widehat{s},\underline{s}) 
 &
 \overline{P_s}& \defeq & E(\overline{s},\widetilde{s},\widehat{s},\underline{s})
\end{array}
\]
%{\color{red}
%\[
%\begin{array}{rcl}
% P(s,\tilde{s},\hat{s},\underline{s}) & \defeq &\sum_{h_j,k_j\in\{0,1\}}
%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^{h_j} \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^{k_j}\ \link{\hat{s}_i}{\hat{s}_o}.P_s\\
%&&\ + \\
%&&
%\sum_{\scriptsize \begin{array}{l}h_j,k_j, \\ h'_j\in\{0,1\}\end{array}}
%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^{h_j} \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^{k_j} \  \startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact} \    (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^{h'_j} \ \link{\underline{s}_i}{\underline{s}_o}.P_s \\
%
%%\sum_{R_1+R_2\geq 0}(\sum_{h,k \in[0,1]} 
%%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^k)^{R_1} \startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact} (\sum_{h ,k\in[0,1]}
%%  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ (\startchain{\tilde{s}_i}\chainedlink{\tilde{s}_o}{\noact}\chainend{\noact})^k)^{R_2} \link{\underline{s}_i}{\underline{s}_o}.P_s\\
%&&\ +\\
%&&\sum_{j\geq 0}(
%  \startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^j\ \ \link{\underline{s}_i}{\underline{s}_o}.\overline{P_s}
%\end{array}
%\]
%}
\[
\begin{array}{rcl}
 E(s,\widetilde{s},\widehat{s},\underline{s}) & \defeq &\sum_{h,k\geq 0}  
  (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ \startchain{\widehat{s}_i}\chainedlink{\widehat{s}_o}{\noact}\chainend{\noact} (\startchain{\widetilde{s}_i}\chainedlink{\widetilde{s}_o}{\noact}\chainend{\noact})^k.P_s\\
&&\ + \\
&&
\sum_{h\geq 0,k\geq 1}
 (\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ \startchain{\underline{s}_i}\chainedlink{\underline{s}_o}{\noact}\chainend{\noact}(\startchain{\widetilde{s}_i}\chainedlink{\widetilde{s}_o}{\noact}\chainend{\noact})^k.P_s\\
&&\ +\\
&&\sum_{h\geq 0}(\startchain{s_i}\chainedlink{s_o}{\noact}\chainend{\noact})^h \ \link{\underline{s}_i}{\underline{s}_o}.\overline{P_s}
\end{array}
\]

The first line of $ E(s,\widetilde{s},\widehat{s},\underline{s})$ accounts for the case where $s$ is tested for presence by $h$ reactions and produced by $k$ reactions, while being provided by the context ($\link{\widehat{s}_i}{\widehat{s}_o}$).
Thus, $s$ will be present at the next step (the continuation is $P_s$). Here $h$ and $k$ are not known a priori and therefore any
combination is possible. 
In practice, by knowing the number of reactions that test $s$, we can bound the maximum 
values of $h$ and $k$.
The second line accounts for the analogous case where $s$ is not provided by the context ($\link{\underline{s}_i}{\underline{s}_o}$). 
The condition $k\geq 1$ guarantees that $s$ will remain present (the continuation is $P_s$).
%Here taking
%$h = \sum_{j\geq0} h_j +h'_j$, $k = (\sum_{j \geq 0} k_j )  + 1$ guarantees that $s$ will be present in the next step. Please note that 
The third line accounts for the case where $s$ is tested for presence, but it is neither produced nor provided by  the context. Therefore, in the next step $s$ will be absent in the system (the continuation is $\overline{P_s}$).
Note that in the case of $\overline{P_s}$ the test for presence of $s$ in the system is just replaced by the test for its absence.
%Process $P_s$ (and $\overline{P_s}$) offers link chain prefixes for all the possible
% permutations of the solid links; nevertheless if the reactions, and their enumeration, is known
% the code of $P_s$ can be drastically simplified.

@@

We exploit names $r_j,p_j$ to join the chains provided by the application of all the reactions. 
Channels $r_j$ and $r_{j+1}$ enclose the enabling/disabling condition of reaction $aj$.
Channels $p_i$ and $p_{j+1}$ enclose the links related to the entities produced by $aj$.
We will see that all the link chain labels of transitions follow the same schema: first we find all the reactions limited to the reactants and inhibitors (chained using $r_j$ channels), then all the supplies by the contexts (chained using the channel $\mathit{cxt}$, to be introduced next), and finally the products for all the reactions (chained using $p_j$ channels). 
In the following there is an example explaining this schema.
%\[
%\begin{array}{lcl@{\qquad}l}
%P_{aj}& \defeq & 
% ^{r_j} \backslash  \blockchain{s_i}{s_o}{s \in R_j} \, \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \in I_j} \, \link{}{}
% \blockchain{\tilde{c}_i}{\tilde{c}_o}{c \in P_j}\, \backslash_{r_{j+1}}.P_{aj} & \mbox{\small  \{$aj$ is applicable\}}\\
%&& + \ \\
%&& \sum_{s\in R_j}\startchain{r_j}\chainedlink{\overline{s}_i}{\noact}\chainedlink{\noact}{\overline{s}_o} \chainend{r_{j+1}}.P_{aj}  & \mbox{\small  \{$aj$ is not applicable\}}\\
%&&+\\
%&& \sum_{e\in I_j}\startchain{r_j}\chainedlink{e_i}{\noact} \chainedlink{\noact}{e_o} \chainend{r_{i+1}}.P_{aj}  & \mbox{\small \{$aj$ is not applicable\}}\\
%\end{array}
%\]

\paragraph{Processes for contexts}
The context can vary at each step $n$ and for each entity $s \in S$, the context must say if the entity is provided or not. Let us denote by $C_n$ the context at step $n$, i.e. the set of entities that are made available. Correspondingly, we introduce another process $\mathit{Cxt}^n$ defined as follows:
\[
\begin{array}{lcl}
\mathit{Cxt}^n &\defeq &
%\left \{
^{cxt} \backslash  \blockchain{s_i}{s_o}{s \in C_n}  \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \not\in C_n}   \backslash_{p_1}. \mathit{Cxt}^{n+1}
%+ \\
% ^{cxt} \backslash  \blockchain{s_i}{s_o}{s \in I2_n} \, \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \in Oj_n} \, \backslash_{p_1}
% \chainedlink{\hat{s}_i}{\noact}\chainedlink{\noact}{\hat{s}_o}\chainend{p_1}. Cxt_{s}^{n+1} & \ \mbox{if the context provides $s$  at the $n$-th step}\\[5pt] 
% \startchain{cxt_j}\chainedlink{\underline{s}_i}{\noact}\chainedlink{\noact}{\underline{s}_o}\chainend{cxt_{j+1}}.Cxt_s^{n+1} & \ \mbox{otherwise}
%\end{array}\right.\\
%Cxt_s & \defeq & Cxt_s^1\\
%\end{array}
%\right.\\
\end{array}
\]

@@

\begin{example}\label{ex:backbone}
 Let ${\cal A}$ be a RS whose specification contains two entities, $s1$ and $s2$, and the reactions $r_1=(s1,\ , s2)$ and $r_2 = (s2, \ , s1)$ that produce $s2$ if $s1$ is present and $s1$ if $s2$ is present. For simplicity we do not consider inhibitors, but the reader can assume a void inhibitor is present in both reactions. Then, we assume an  extended interactive process $\pi=(\gamma,\delta)$ where the context $\gamma$ provides  $s1$ and $s2$ at every step, but we assume that only $s1$ is initially present. The corresponding c\CNA \ process is $\llbracket{\cal A},\gamma\rrbracket \defeq \restrict{\mathit{names}} \llparenthesis{\cal A},\gamma\rrparenthesis$, with 
$$
\llparenthesis{\cal A},\gamma\rrparenthesis \defeq 
I ~|~ P_{s1}~|~\overline{P}_{s2}~|~ P_{r1}~|~P_{r2}~|~\mathit{Cxt}
$$
 where:
 \[
 \begin{array}{rcl}
 P_{r_1} &\defeq& \startchain{r_1}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}\chainedlink{r_2}{\noact} \chainedlink{\noact}{p_1}\chainedlink{\widetilde{s2}_i}{\noact}\chainedlink{\noact}{\widetilde{s2}_o}\chainend{p_2}.P_{r_1} + \dots;   \\[6pt]
  P_{r_2} &\defeq& \dots + \startchain{r_2}\chainedlink{\overline{s2}_i}{\noact}\chainedlink{\noact}{\overline{s2}_o}\chainedlink{\mathit{cxt}}{\noact} \chainedlink{\noact}{p_2}\chainend{\silent}.P_{r_2} + \dots; \\[6pt]
   P_{s1} &\defeq& \startchain{s1_i}\chainedlink{s1_o}{\noact}\chainedlink{\noact}{\widehat{s1}_i}\chainend{\widehat{s1}_o}
 %{\noact}\chainedlink{\noact}{\tilde{s1}_i}\chainedlink{\tilde{s1}_o}
 .P_{s1} + \dots; \\[6pt]
  \overline{P}_{s2} &\defeq& \startchain{\overline{s2}_i}\chainedlink{\overline{s2}_o}{\noact}
   \chainedlink{\noact} {\widehat{s2}_i}\chainedlink{\widehat{s2}_o}{\noact}\chainedlink{\noact} {\widetilde{s2}_i}\chainend{\widetilde{s2}_o}.P_{s2} + \dots;\\[6pt]
 \mathit{Cxt}& \defeq & \startchain{\mathit{cxt}}\chainedlink{\widehat{s1}_i}{\noact}\chainedlink{\noact}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\noact}\chainedlink{\noact}{\widehat{s2}_o}\chainend{p_1}.\mathit{Cxt}
  %Cxt& \defeq & \startchain{cxt}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}
  %\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}.Cxt
 %Cxt_{s1}& \defeq & \startchain{cxt_1}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}\chainend{cxt_2}.Cxt_{s1} &
 % Cxt _{s2} &\defeq& \startchain{cxt_2}\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}. Cxt_{s2}
\end{array}
\]

For clarity of exposition, we show the code of the processes just in part, to focus on the prefixes that will be involved in the first transition of the system. In Figure~\ref{fig:backbone} we show the  structure of a link chain label related to the execution of such a transition. The yellow blocks are referred to init process $I$, to the processes encoding the reactions, $P_{r_1}$ and $P_{r_2}$, and to the context $\mathit{Cxt}$. As the figure puts in evidence, these two kinds of processes determine the structure of the link chain, from end to end, i.e. from the left $\silent$ to the right one. We could say that these processes form the \emph{backbone} of the interaction.
   In contrast, the processes encoding the entities, $P_{s1}$,  $\overline{P}_{s2}$, provide the
    solid links to be merged with the virtual links of the backbone (i.e. to be plugged in the backbone).
 Formally, we have
 $$
 \llparenthesis{\cal A},\gamma\rrparenthesis \xrightarrow{
 \startchain{\silent}
 \chainedlink{r_1}{r_1}
 \chainedlink{s1_i}{s1_i}
 \chainedlink{s1_o}{s1_o}
 \chainedlink{r_2}{r_2}
 \chainedlink{\overline{s2}_i}{\overline{s2}_i}
 \chainedlink{\overline{s2}_o}{\overline{s2}_o}
 \chainedlink{\mathit{cxt}}{\mathit{cxt}}
 \chainedlink{\widehat{s1}_i}{\widehat{s1}_i}
 \chainedlink{\widehat{s1}_o}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\widehat{s2}_i}
  \chainedlink{\widehat{s2}_o}{\widehat{s2}_o}
  \chainedlink{p_1}{p_1}
  \chainedlink{\widetilde{s2}_i}{\widetilde{s2}_i}
  \chainedlink{\widetilde{s2}_o}{\widetilde{s2}_o}
  \chainedlink{p_2}{p_2}
  \chainend{\silent}
 }
 P'
 $$

@@

\paragraph{The flat function}
For technical reasons, our transition labels are quite verbose; then, to simplify their processing, we introduce a function that takes a solid link chain and returns a simple string by eliminating all the channel matching pairs leaving just one placeholder for them. This transformation is harmless, in the sense that it retains all the information in the chain, because it is applied to complete chains only. 
The function $\mathit{flat}(\cdot)$ is defined inductively as follows:
%\[
%\begin{array}{lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl}
%\mathit{flat}(\epsilon)&\defeq & \epsilon &
%\mathit{flat}(\link{\gamma}{\gamma'}) &  \defeq & \left \{ \begin{array}{cl} \epsilon & \mathrm{if } (\gamma = \beta_i \wedge \gamma = \beta_o)  \vee \gamma = \silent\\
%\gamma' & \mathrm{otw }  \end{array} \right. &
%%\mathit{flat}(\link{\alpha}{\silent})& \defeq & \epsilon &
%%\mathit{flat}(\link{\alpha}{\beta}) & \defeq &\beta \\
%\end{array}\]

@@

% \begin{proposition}[correctness2]
% \label{prop:correctness1}
% Let $P$ be a c\CNA \ process such that $\exists$ a $rs$ ${\cal A}$ and a $n$-step interactive process $\pi=(\gamma,\delta)$ with $\gamma =C_0,\dots,C_n$ and $\delta = D_0,\dots,D_n$ such that $P= [|{\cal A},\gamma|]$. 
% %We set $\pi'=(\gamma',\delta')$ with $\gamma'=C_0',\dots,C_{n-1}'$,  $\delta'=D_0',\dots,D_{n-1}$,  where $C_0'= C_1 \cup D_1$, $C_i'= C_{i+1}$, $D_i=D_{i+1}'$,  and $1 \leq i < n$.
%Then,   $\exists$ $P'$ and $\pi'=(\gamma',\delta')$ such that $P \xrightarrow{\link{\silent}{\silent}}P'$, $P' =[|{\cal A},\gamma'|]$.
% \end{proposition}

Now, we extend the previous result to a series of transitions.
